---
title: "Lab5HWPatelDhiraj"
author: "Dhiraj Patel"
date: "5/4/2018"
output: html_document
---
### Lab 5 Homework

**Due Tuesday, 9 May 2017 at 11:59pm.**

Human names are often of obscure origin, but many have fairly obvious sources. For example, "Johnson" means "son of John," "Smith" refers to an occupation, and "Whitlock" means "white-haired" (from "white locks"). In Lancashire, U.K., a fair number of people are named "Shufflebottom," a name whose origins remain obscure. 

Before children learn to walk, they move around in a variety of ways, with most infants preferring a particular mode of transportation. Some crawl on hands and knees, some belly crawl commando-style, and some shuffle around on their bottoms. A group of researchers decided to ask whether the name "Shufflebottom" might be associated with a propensity to bottom-shuffle. To test this, they compared the frequency of bottom-shufflers among infants with the last name "Shufflebottom" to the frequency for infants named "Walker." (By the way, this study is real. See Fox et al. 2002.)

They found that 11 out of 43 Walkers moved by bottom-shuffling, while 12 out of 45 Shufflebottoms did. Is there a significant difference between the groups?

<span style="color:green">
1.	State the null hypothesis and the alternative hypothesis for the independence test.
</span>

<span style="color:blue">Null hypothesis: There is no association between the name 'Shufflebottom' and a propensity to bottom shuffle. Alternative: There is an association between the name 'Shufflebottom' and a propensity to bottom-shuffle.</span>

<span style="color:green">
2.	Enter the observed values into a matrix in R. Next calculate the expected counts based on the assumption that hypothesis is true. Perform these calculations by hand and enter them into a matrix in R.
</span>

```{r}
ocVec<-c(11,32,12,33)
walkData <- matrix(data=ocVec, ncol=2,nrow=2,byrow=TRUE)

```


<span style="color:green">
3.	Calculate the chi-squared statistic using the function you wrote in lab.
</span>

```{r}
#Function Written in Lab5: 
chiSquareTest <- function(observed,expected){
  count<-0
  for(i in 1:nrow(observed)){
    for(j in 1:ncol(observed)){
      count <- count + (((observed[i,j]-expected[i,j])^2)/(expected[i,j]))
    }
  }
  return(count)
}
expWalkDataVec <- c(11.2,11.8,31.8,33.2)
expWalkData<-matrix(data=expWalkDataVec,nrow=2,ncol=2,byrow=TRUE)
chiSquareTest(walkData,expWalkData)
```


<span style="color:green">
4.	Find the threshold chi-squared value using the `qchisq()` function with the correct degrees of freedom and significance levels $\alpha = 0.05, 0.01, 0.001$ and decide whether to reject the independence hypothesis at each significance level. 

```{r}
qchisq(0.95,1) #we reject the null hypothesis because the P-value is greater than the threshold. 
qchisq(0.99,1) #we reject the null hypothesis because the P-value is greater than the threshold
qchisq(0.999,1) #we reject the null hypothesis because the P-value is greater than the threshold. 
```

</span>